Graph Topological Aspects of Granger Causal Network Learning

### 2019-11-16 paper_arxiv/supplementary_material.tex
arxiv version.

\documentclass{statsoc}
\usepackage{fullpage}
\usepackage{ccfonts,eulervm}
\usepackage{tabularx}
\usepackage{xr}
\usepackage{chngcntr}
\usepackage{appendix}
\newcounter{lemma}
\newcounter{definition}
\counterwithin{lemma}{section}
\counterwithin{definition}{section}
\counterwithin{equation}{section}
\externaldocument{graph_topology_granger_causality}

\usepackage{geometry}

\geometry{
  textwidth=33pc,
  textheight=\dimexpr48\baselineskip+\topskip\relax,
  marginparsep=11pt,
  marginparwidth=107pt,
  footnotesep=6.65pt,
  headheight=9pt,
  headsep=9pt,
  footskip=30pt,
}
\bibliographystyle{chicago}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\usepackage{multirow}
\usepackage{booktabs}

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\MA{\mathsf{MA}}  % MA(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things
\newcommand{\gcgpath}[2]{#1 \rightarrow \cdots \rightarrow #2}  % A shorter path command

% \usepackage{fullpage}
\usepackage{framed}

% Figures
\usepackage{graphicx}
\usepackage{caption}  % This is not recommended?
\usepackage{subcaption}
% \usepackage{wrapfig}
% \usepackage{svg}

% Math packages, theorem definitions and numbering
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
% \usepackage{mathrsfs} % Fancy scripted font
\usepackage{bm}  % Bold math

% Misc packages
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algorithm2e} %{algorithm} environment
\usepackage{soul}  % \hl highlighting
\usepackage{color}
\usepackage{mathtools}  % For my \ceil function

% Theorems (with italics)
% \theoremstyle{plain}  % Style definition removes italics
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

% \theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumptions}{Assumptions}

% keywords
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

% General
\def\defeq{\overset{\Delta}{=}}  % Equal with triangle
\def\cl{\mathsf{cl\ }}  % Closure
\newcommand{\sgn}[1]{\mathsf{sgn}(#1)}  % sign function

% Calculus
\def\d{\mathsf{d}}  % Differential operator

% Functions
\def\ln{\mathsf{ln\ }}  % Natural logarithm
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}  % Ceiling

% Probability
\def\H{\mathcal{H}}  % Hilbert space
\def\E{\mathbb{E}}  % Expectation
\def\Var{\text{Var}}  % Variance
\def\P{\mathbb{P}}  % Probability Measure
\def\F{\mathcal{F}}  % A sigma algebra
\def\sX{\mathcal{X}}  % Another sigma algebra
\def\KL{\mathbf{D}_{KL}}  % KL divergence
\def\bF{\mathbf{F}}  % Whole F-meas space

% Standard sets
\def\Z{\mathbb{Z}}  % Set of integers
\def\R{\mathbb{R}}  % Set of real numbers
\def\C{\mathbb{C}}  % Set of complex numbers
\def\N{\mathbb{N}}  % Set of natural numbers
\def\ball{\mathbb{B}}  % Open ball
\def\clball{\overline{\ball}}  % Closed ball

% Linear algebra
\def\rk{\mathsf{rk }}  % The rank
\def\tr{\mathsf{tr }}  % The trace
\def\T{\mathsf{T}}  % Transpose notation
\def\c{\mathsf{c}}  % complement
\def\dg{\mathsf{dg }}   %  Diagonal vector of a matrix
\def\Dg{\mathsf{Dg }}   %  Diagonal matrix from a vector
\def\ind{\mathbf{1}}  % Ones vector or indicator
\def\matvec{\textbf{vec}}  % Vector operator
\def\<{\langle}  % < Inner product
\def\>{\rangle}  % > Inner product
\newcommand{\inner}[2]{\langle #1, #2 \rangle}  % Inner product
\newcommand{\innerT}[2]{#1^\T #2}  % Inner product for finite vectors

\graphicspath{{../figures/}}

\title{Graph Topological Aspects of Granger Causal Network Learning\\
  \large Supplementary Material}

\author[Author 1 {\it et al.}]{R. J. Kinnear}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}
\email{Ryan@Kinnear.ca}

\author{R. R. Mazumdar}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}

\begin{document}

\appendix

@@

\begin{proof}
  % Geweke uses the log of the ratio of the determinants of the residual variances
  % 
  % The equivalence $(1) \iff (3)$ is essentially a restatement of the
  % definition, but is in line with the seminal work of Geweke
  % \cite{geweke1982measurement}, \cite{geweke1984}.

  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{(-j)}})$ for $y \in H_t^{(i)}$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^{(i)} \subseteq \H_{t - 1}^{(-j)}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(-j)}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$.  The result follows since
  $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result of the equivalence
  in Definition \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

% I believe this is an incorrect proof since it relies on inverting
% (1 - B_{jj}(z)), which need not be invertible.  For example,
% B(z) = [[r, -a], [a, 0]]z^{-1} is invertible whenever
% max(|r - root(r^2 - 4a^2)|, |r + root(r^2 - 4a^2)|) < 2
% which includes values of r > 1.

% \begin{proof}
%   By Theorem \ref{thm:granger_causality_equivalences} it suffices to show that

%   \begin{equation*}
%     \forall \psi \in \H_{t - 1}^j\ \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{\psi - \linE{\psi}{\H_{t - 1}^i}} = 0.
%   \end{equation*}

%   which by the orthogonality principle and by representing
%   $\psi \in \H_{t - 1}^j$ via the action of some strictly causal filter
%   $\Phi(z)$ on $x_j(t)$ is equivalent to

%   \begin{equation}
%     \label{eqn:proof_inner0}
%     \inner{x_i(t)}{\Phi(z)x_j(t) - \linE{\Phi(z)x_j(t)}{\H_{t - 1}^i}} = 0.
%   \end{equation}

%   Given a node $u$, consider the partition of $\pa{u}$:

%   \begin{align*}
%     C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, k \ne i \}\\
%     C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i \}.
%   \end{align*}

%   We can then expand the parents of $j$ using Equation \ref{eqn:parent_expansion} as

%   \begin{align*}
%     x_j(t) &= v_j(t) + \B_{jj}(z)x_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t)\\
%     &= \widetilde{A}_{jj}(z)\Big(v_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t)\Big),
%   \end{align*}

%   % \todo{Is this true??}
%   where $(1 - B_{jj}(z))^{-1} = \widetilde{A}_{jj}(z)$, which exists
%   due to the invertibility of $\A(z)$. When this expression is
%   substituted into the left hand side of Equation
%   \ref{eqn:proof_inner0} it results in

%   \begin{equation}
%     \label{eqn:proof_inner1}
%     \begin{aligned}
%       \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{A}_{jj}(z)\Phi(z)B_{jk}(z) x_k(t) &- \linE{\widetilde{A}_{jj}(z)\Phi(z)B_{jk}(z) x_k(t)}{\H_{t - 1}^i}\big)},
%     \end{aligned}
%   \end{equation}

%   where (by Proposition \ref{prop:separated_ancestor_uncorrelated}) we
%   have eliminated each of the terms from $C_0(u)$, and by Lemma
%   \ref{lem:vj_perp} the terms involving $v_j(t)$.

%   Now, for any $u \in \anc{j}$, we must have $u \not \in \anc{i}$, since
%   otherwise $u \in \anc{i} \cap \anc{j}$, which is empty by hypothesis.
%   Moreover, for any $u \in \anc{j}$, we have
%   $\anc{i} \cap \anc{u} = \emptyset$, since
%   $w \in \anc{i} \cap \anc{u} \implies w \in \anc{i} \cap \anc{j}$. Therefore,
%   we can continue the above process recursively splitting each
%   $k \in C_1(j)$ into $C_0(k)$ and $C_1(k)$) upwards along each path
%   from $j$ to $i$, terminating whenever we encounter $i$.  This process
%   is guaranteed to terminate since $i \in \anc{k}$.  This leaves us with
%   some strictly causal filter $\Psi(z)$ such that Expression
%   \ref{eqn:proof_inner1} is equivalent to

%   \begin{equation}
%     \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}},
%   \end{equation}

%   which is $0$ since $\Psi(z)x_i(t) \in \H_{t - 1}^i$ and
%   therefore $\linE{\Psi(z)x_i(t)}{\H_{t - 1}^i} = \Psi(z)x_i(t)$.
% \end{proof}

@@

\begin{lemma}[Expansion in SCCs]
  \label{lem:scc_expansion}
  Given some $j \in [n]$, the process $x_{S(j)}(t)$ can be represented
  by
% {\substack{k \in \pa{a_0} \\ k \ne a_1}}
  \begin{equation}
    \label{eqn:scc_parent_expansion}
    x_{S(j)}(t) = v_{S(j)}(t) + \B_{S(j)}(z)x_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)},
  \end{equation}

@@

  and this is $0$ since $\Phi(z)F(z)x_i(t) \in \H_{t - 1}^{(i)}$.

  % Using the above partition of $\pa{j}$, we can expand $x_j(t)$ in
  % terms of it's parents, and recursively expand nodes in $C_1$
  % (cancelling all nodes in $C_0$, and recalling that
  % $C_2 = \emptyset$) until we reach $i$.

  % Using this partition, we will expand $x_j(t)$ in terms of it's
  % parents, and recursively expand nodes in $C_1$ until reaching $i$.
  % For the first step equation \eqref{eqn:parent_expansion} gives us:

  % \begin{equation}
  %   \label{eqn:xj_partition_expansion}
  %   x_j(t) = \widetilde{\A}_{jj}(z) \Big(v_j(t) + \sum_{k \in C_0(j)} \B_{ik}(z)x_k(t) + \sum_{k \in C_1(j)}\B_{ik}(z)x_k(t)\Big).
  % \end{equation}

  % Substituting (\ref{eqn:xj_partition_expansion}) into
  % (\ref{eqn:sufficient_inner_prod}) we have cancellation of all terms
  % in which $k \in C_0(j)$ by Proposition \ref{prop:ancestor_properties},
  % and cancellation of the term involving $v_j(t)$ by Lemma \ref{lem:vj_perp}.

  % We are therefore left with

  % \begin{equation}
  %   \label{eqn:refwornv}  % An arbitrary code for reference in just one place
  %   \inner{x_i(t)}{\sum_{k \in C_1(j)} \big(\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t) - \linE{\widetilde{\A}_{jj}(z)\Phi(z) \B_{ik}(z)x_k(t)}{\H_{t - 1}^i}\big)}.
  % \end{equation}

  % Since each such $k$ inherits the needed properties $(a), (b)$ we can
  % again (as in Proposition \ref{prop:ancestor_properties}) recursively
  % expand this expression until we reach $i$, obtaining a strictly
  % causal $\Psi(z)$ such that Expression \eqref{eqn:refwornv} is equal to

  % \begin{equation}
  %   \inner{x_i(t)}{\Psi(z)x_i(t) - \linE{\Psi(z)x_i(t)}{\H_{t - 1}^i}} = 0.
  % \end{equation}
\end{proof}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
  which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
  $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
  $i \pwgc j$.  Moreover, if Assumption \ref{ass:T_causality} is
  satisfied, then $j \pwgc i \iff i \pwgc j$.
  % Suppose that Assumption \ref{ass:T_causality} is satisfied.  Then if
  % there exists a $k$ which confounds $(i, j)$ we have
  % $i \pwgc j \implies j \pwgc i$.  Moreover, if $T_{ij}(z)$ is not
  % constant, then $i \pwgc j$.
\end{proposition}

% QUESTION: Does $\gcg$ need to be a strongly causal DAG for this proposition?

% \begin{proposition}
%   \label{prop:persistence_converse}
%   Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is
%   persistent, then if there exists a $k$ which confounds $(i, j)$ we
%   have $i \pwgc j \implies j \pwgc i$.
% \end{proposition}
\begin{proof}
  Recalling Theorem \ref{thm:granger_causality_equivalences}, consider
  some $\psi \in \H_{t - 1}^{(j)}$ and represent it as
  $\psi(t) = F(z)x_j(t)$ for some strictly causal filter $F(z)$.
  Then

@@

% , $(c)$ is obtained through cancellations from Lemma \ref{lem:vj_perp}, $(d)$ follows since $\H_i$ is self-adjoint, $(e)$ follows by definition of $\H_i$: $\exists H_i(z)$ such that $$H_i(z)x_i(t) = \H_i \sum_{\ell \in \anc{i}}\A_{i\ell}v_\ell(t),$$ $(f)$ follows by again expanding $x_i(t)$ with Equation \eqref{eqn:ancestor_expansion}, and finally $(g)$ by cancelling orthogonal components of $v(t)$.

@@

% This proof is not correct -- there can be cancellation
% \begin{proof}
%   We will show that $j \pwgc i$, the other being symmetric.  First
%   note that by the contrapositive of proposition
%   \ref{prop:sc_graph_common_anc} we cannot have $i \in \anc{j}$ or
%   $j \in \anc{i}$ (else $k$ would not be a confounder) and therefore
%   \textit{every} $u \in \anc{i}\cap\anc{j}$ will be a confounder.

%   It is sufficient to show that $\exists \psi \in \H_{t - 1}^{(j)}$
%   such that

%   \begin{equation*}
%     \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}} \ne 0.
%   \end{equation*}

%   To this end, let $F(z)$ be an arbitrary but strictly causal linear
%   filter.  We apply equation \eqref{eqn:ancestor_expansion} to
%   $x_i(t)$ and $\psi \defeq F(z)x_j(t)$:

%   \begin{align*}
%     &\inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(i)}}}\\
%     &\overset{(a)}{=} \inner{\psi - \linE{\psi}{\H_{t - 1}^{(i)}}}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
%     &\overset{(b)}{=} \inner{\sum_{k \in \anc{j}}\big(F(z)\A_{jk}(z)v_k(t) \\&- \linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}\big)}{\A_{ii}(z)v_i(t) + \sum_{k \in \anc{i}}\A_{ik}(z)v_k(t)}\\
%     &\overset{(c)}{=} \sum_{k \in \anc{i}\cap\anc{j}}\Big(\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} \\&- \inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{ii}v_i(t)}\\
%     &- \sum_{\ell \in \anc{i}}\inner{\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}}}{\A_{i\ell}(z)v_\ell(t)}\Big)
%   \end{align*}

%   where in $(a)$ we have removed the $\linE{x_i(t)}{\H_{t - 1}^{(i)}}$
%   term via the orthogonality principle, in $(b)$ there is no
%   $F(z)\A_{jj}(z)v_j(t)$ term since due to $j \not\in \anc{i}$ it is
%   orthogonal to $\H_t^{(i)}$.  Finally, $(c)$ follows by applying
%   orthogonality properties of $v(t)$, as well as the fact that
%   $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ for
%   $k \not \in \anc{i}$.  Note that
%   $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} \in \H_{t - 1}^{(i)}$
%   and therefore there is in general no cancellation in the final term
%   above for $\ell \in \anc{i}$.

%   \todo{This is clearly not immediately evident}

%   This is $0$ for every $F$ if and only if for all $F$ and
%   $\forall k \in \anc{i} \cap \anc{j}$ we have
%   $$\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$$ and
%   $$\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0,$$ which by
%   Corollary \ref{cor:time_lag_cancellation} occurs if and only if
%   $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible since by persistence
%   $\tau_0(\A_{jk}) < \infty$ and  $\tau_\infty(\A_{ik}) = \infty$.
% \end{proof}

@@

  % \BlankLine
  \caption{Pairwise Granger Causality Algorithm (PWGC)}
  \label{alg:pwgr}
  % \TitleOfAlgo{Pairwise Graph Recovery (PWGC)}
  % \Input{Pairwise Granger causality relations between a persistent
  % process of dimension $n$ whose joint Granger causality
  % relations are known to form a strongly causal DAG $\gcg$.}
  \Input{Pairwise Granger causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
    % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parentless nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

    $D_{k0} \leftarrow \emptyset$\\
    \For{$r = 1, \ldots, k$} 
    {
      $Q \leftarrow E_{k - 1} \cup \big(\bigcup_{\ell = 0}^{r - 1} D_{k\ell}\big)$ \texttt{\# currently known edges}\\
      $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times P_k\ |\ (i, j) \in W,\ \text{no } i \rightarrow j \text{ path in } Q\}$
      % $E_k \leftarrow E_{k - 1} \cup D_{kr}$ \texttt{\# add edges to }$E_k$ \label{alg:inner_loop_end}\\
      % $W_{k + 1} \leftarrow W_k \setminus D_{kr}$  \texttt{\# remove edges from consideration}\\
    }
    $E_k \leftarrow E_{k - 1} \cup \big(\bigcup_{r = 1}^k D_{kr}\big)$ \texttt{\# update } $E_k$ \texttt{ with new edges}\\
    % \label{alg_line:inner_loop_end}\\
    % $W_{k + 1} \leftarrow W_k \setminus \big(\bigcup_{r = 0}^k D_{kr}\big)$  \texttt{\# remove edges from consideration}\\
    $k \leftarrow k + 1$
  }
  \Return{$E_{k - 1}$}
\end{algorithm}

@@

    \While{$S \ne \emptyset$}{
      $S \leftarrow S \setminus P_{k - 1}$ \texttt{\# remove processed nodes}\\
      % $\mathcal{I}_k \leftarrow \big(\sum_{j \in S: (j, i) \in W_\delta} F_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# intra-}$S$ \texttt{incident strength}\\
      $\mathcal{I}_k \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$\\
      $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) < \ceil{\text{min}(\mathcal{I}_k)}\}$\\
      \If{$P_k = \emptyset$}{
        $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) \le \ceil{\text{min}(\mathcal{I}_k)}\}$
      }
      \;
      \texttt{\# add strongest edges, maintaining strong causality}\\
      $U_k \leftarrow \bigcup_{r = 1}^k P_{k - r}$ \texttt{\# Include all forward edges}\\
      \For{$(i, j) \in \mathsf{sort}\Big(\{(i, j) \in U_k \times P_k\ |\ (i, j) \in W_\delta\} \mathsf{\ by\ descending\ } F_{ji}\Big)$} {
        \If{$\mathsf{is\_strongly\_causal}(E \cup \{(i, j)\})$} {
          \texttt{\# }$\mathsf{is\_strongly\_causal}$ \texttt{can be implemented by keeping}\\
          \texttt{\# track of ancestor / descendant relationships}\\
          $E \leftarrow E \cup \{(i, j)\}$
        }
      }
      $k \leftarrow k + 1$\\
    }
    \Return{$([n], E)$}
\end{algorithm}

@@

\clearpage
\bibliography{paper_bib}
\end{document}

>>>>

% Appendix with proofs and more figures

@@

\begin{proof}
  $(a) \Rightarrow (b)$ follows as a result of the uniqueness of orthogonal
  projection (i.e. the best estimate is necessarily the coefficients
  of the model).  $(b) \Rightarrow (c)$ follows since in computing
  $(y - \linE{y}{\H_{t - 1}^{(-j)}})$ for $y \in H_t^{(i)}$ it is sufficient
  to consider $y = x_i(t)$ by linearity, then since
  $H_{t - 1}^{(i)} \subseteq \H_{t - 1}^{(-j)}$ we have
  $(x_i(t) - \linE{x_i(t)}{\H_{t - 1}^{(-j)}}) = v_i(t)$ since
  $\B_{ij}(z) = 0$.  The result follows since
  $v_i(t) \perp \H_{t - 1}$.  $(c) \iff (d)$ is a result of the equivalence
  in Definition \ref{lem:conditional_orthogonality_equivalence}.  And,
  $(d) \implies (a)$ follows directly from the Definition.
\end{proof}

@@

@@

\begin{lemma}[Expansion in SCCs]
  \label{lem:scc_expansion}
  Given some $j \in [n]$, the process $x_{S(j)}(t)$ can be represented
  by

  \begin{equation}
    \label{eqn:scc_parent_expansion}
    x_{S(j)}(t) = v_{S(j)}(t) + \B_{S(j)}(z)x_{S(j)}(t) + \sum_{\substack{s \in S(j) \\ k \in \pa{s} \cap S(j)^\c}} B_{sk}(z)x_k(t)e_s^{S(j)},
  \end{equation}

@@

  and this is $0$ since $\Phi(z)F(z)x_i(t) \in \H_{t - 1}^{(i)}$.
\end{proof}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Fix $i, j \in [n]$ and suppose $\exists k \in \anc{i} \cap \anc{j}$
  which confounds $i, j$.  Then, if $T_{ij}(z)$ is not causal we have
  $j \pwgc i$, and if $T_{ij}(z)$ is not anti-causal we have
  $i \pwgc j$.  Moreover, if Assumption \ref{ass:T_causality} is
  satisfied, then $j \pwgc i \iff i \pwgc j$.
\end{proposition}

\begin{proof}
  Recalling Theorem \ref{thm:granger_causality_equivalences}, consider
  some $\psi \in \H_{t - 1}^{(j)}$ and represent it as
  $\psi(t) = F(z)x_j(t)$ for some strictly causal filter $F(z)$.
  Then

@@

@@

@@

  \caption{Pairwise Granger Causality Algorithm (PWGC)}
  \label{alg:pwgr}
  \Input{Pairwise Granger causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\

    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parentless nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

    $D_{k0} \leftarrow \emptyset$\\
    \For{$r = 1, \ldots, k$} 
    {
      $Q \leftarrow E_{k - 1} \cup \big(\bigcup_{\ell = 0}^{r - 1} D_{k\ell}\big)$ \texttt{\# currently known edges}\\
      $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times P_k\ |\ (i, j) \in W,\ \text{no } i \rightarrow j \text{ path in } Q\}$
    }
    $E_k \leftarrow E_{k - 1} \cup \big(\bigcup_{r = 1}^k D_{kr}\big)$ \texttt{\# update } $E_k$ \texttt{ with new edges}\\
    $k \leftarrow k + 1$
  }
  \Return{$E_{k - 1}$}
\end{algorithm}

@@

    \While{$S \ne \emptyset$}{
      $S \leftarrow S \setminus P_{k - 1}$ \texttt{\# remove processed nodes}\\
      $\mathcal{I}_k \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$\\
      $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) < \ceil{\text{min}(\mathcal{I}_k)}\}$\\
      \If{$P_k = \emptyset$}{
        $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) \le \ceil{\text{min}(\mathcal{I}_k)}\}$
      }
      \;
      \texttt{\# add strongest edges, maintaining strong causality}\\
      $U_k \leftarrow \bigcup_{r = 1}^k P_{k - r}$ \texttt{\# Include all forward edges}\\
      \For{$(i, j) \in \mathsf{sort}\Big(\{(i, j) \in U_k \times P_k\ |\ (i, j) \in W_\delta\} \mathsf{\ by\ descending\ } F_{ji}\Big)$} {
        \If{$\mathsf{is\_strongly\_causal}(E \cup \{(i, j)\})$} {
          \texttt{\# }$\mathsf{is\_strongly\_causal}$ \texttt{can be implemented by keeping}\\
          \texttt{\# track of ancestor / descendant relationships}\\
          $E \leftarrow E \cup \{(i, j)\}$
        }
      }
      $k \leftarrow k + 1$\\
    }
    \Return{$([n], E)$}
\end{algorithm}

@@

\clearpage

### 2019-11-16 presentations/student_dynamical_systems_waterloo_2019/pres.tex
arxiv version.

>>>>

\begin{frame}{Qualitative Structure of Granger Causality Graphs}
  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_rotation_shape.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_star_shaped.pdf}\pause
    \end{column}
  \end{columns}

  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_loop_shaped.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_diamond_shaped.pdf}
    \end{column}
  \end{columns}
\end{frame}
